Long-term care homes were hit hardest by 2023 disease outbreaks in Toronto healthcare facilities

### 2024-01-25 outputs/paper/paper.qmd
Adding a Data Cleaning subsection to the Appendix section of the paper
Removed a space
Added data labels to the graphs and removed the tables

---
title: "Long-term care homes were hit hardest by 2023 disease outbreaks in Toronto healthcare facilities"
author: 
  - Benny Rochwerg
thanks: "A GitHub Repository containing all data, R code, and other files used in this investigation is located here: <https://github.com/bennyrochwerg/healthcare-outbreaks-toronto>"
date: 01-23-2024
date-format: long
abstract: "In this report, 2023 Toronto Public Health data on outbreaks in healthcare facilities was analyzed. Long-term care homes comprised a majority of the outbreaks, approximately 95% of the outbreaks were respiratory, and almost two-thirds of the outbreaks had COVID-19 as their first known cause. This highlights the need for enhanced public health measures in healthcare facilities, particularly in long-term care homes. Future research should assess disease outbreaks in additional settings, such as workplaces and schools, to gain further insight into the extent of this issue."
format: pdf
number-sections: true
# The toc parameter used to create a table of contents was obtained from
# Quarto documentation:
# https://quarto.org/docs/reference/formats/pdf.html
toc: true
# The tbl-colwidths parameter used to adjust the table column widths was
# obtained from Quarto documentation:
# https://quarto.org/docs/authoring/tables.html
tbl-colwidths: [50, 20, 20]
bibliography: references.bib
---

@@

# install.packages("tidyverse")
# install.packages("here")
# install.packages("knitr")
library(tidyverse)
library(here)
library(knitr)
```

@@

To do this, 2023 Toronto Public Health data on outbreaks in healthcare facilities was first obtained as described in @sec-data-overview. Based on this dataset, it was discovered that long-term care homes comprised a majority of the outbreaks, that approximately 95\% of the outbreaks were respiratory, and that almost two-thirds of the outbreaks had COVID-19 as their first known cause (@sec-data-results). As discussed in @sec-data-discussion, the high prevalence of outbreaks in long-term care homes has likely been exacerbated by the removal of a universal masking requirement, suggesting that the Government of Ontario should consider reinstating this policy to protect these vulnerable individuals. With respect to the structure of this paper, @sec-data includes an overview of the data, results, and discussion; @sec-acknowledgements acknowledges a valuable source of knowledge in creating this report; and @sec-appendix includes supplementary information.

@@

After loading the dataset using the R programming language [@citeR] and the `here` package [@citehere], the `tidyverse` [@citetidyverse] and `knitr` [@citeknitr] packages were used to generate tables and graphs. In doing so, R code was adapted from @tellingstorieswithdata.

@@

```{r}
#| echo: false
#| warning: false
#| label: tbl-outbreaks-by-location-type
#| tbl-cap: "Outbreaks by type of healthcare location in Toronto in 2023"

#### First Table ####

@@

outbreak_cleaned_data |>
  count(`Type of Location`) |>
  rename(`Type of location` = `Type of Location`,
         `Number of outbreaks` = n) |>
  mutate(`Proportion of total outbreaks` =
           round(`Number of outbreaks` / nrow(outbreak_cleaned_data), 2)) |>
  arrange(desc(`Number of outbreaks`)) |>
  kable()
```

@@

outbreak_cleaned_data |>
  count(`Type of Location`) |>
  # The reorder function used for sorting the columns in descending order was obtained
  # from Stack Overflow:
  # https://stackoverflow.com/questions/25664007/reorder-bars-in-geom-bar-ggplot2-by-value
  ggplot(mapping = aes(x = reorder(`Type of Location`, n), y = n)) +
  geom_col() +
  theme_bw() +
  labs(x = "Type of location", y = "Number of outbreaks") +
  coord_flip()
```

@tbl-outbreaks-by-location-type and @fig-outbreaks-by-location-type illustrate that long-term care homes comprised a majority (approximately 56\%) of the total number of 2023 healthcare outbreaks in Toronto. This was followed by retirement homes (approximately 18\%), chronic care settings in hospitals (approximately 12\%), acute care settings in hospitals (approximately 10\%), psychiatric care settings in hospitals (approximately 2\%), and transitional care locations (approximately 1\%). Note that these percentages do not add up to 100\% as the proportions in @tbl-outbreaks-by-location-type were rounded to two decimal places.

@@

```{r}
#| echo: false
#| warning: false
#| label: tbl-outbreaks-by-outbreak-type
#| tbl-cap: "Outbreaks by type of outbreak for Toronto healthcare locations in 2023"

#### Second Table ####

@@

outbreak_cleaned_data |>
  count(`Type of Outbreak`) |>
  rename(`Type of outbreak` = `Type of Outbreak`,
         `Number of outbreaks` = n) |>
  mutate(`Proportion of total outbreaks` =
           round(`Number of outbreaks` / nrow(outbreak_cleaned_data), 2)) |>
  arrange(desc(`Number of outbreaks`)) |>
  kable()
```

@@

outbreak_cleaned_data |>
  count(`Type of Outbreak`) |>
  # The reorder function used for sorting the columns in descending order was obtained
  # from Stack Overflow:
  # https://stackoverflow.com/questions/25664007/reorder-bars-in-geom-bar-ggplot2-by-value
  ggplot(mapping = aes(x = reorder(`Type of Outbreak`, -n), y = n)) +
  geom_col() +
  theme_bw() +
  labs(x = "Type of outbreak", y = "Number of outbreaks")
```

@tbl-outbreaks-by-outbreak-type and @fig-outbreaks-by-outbreak-type show that the vast majority of 2023 healthcare outbreaks in Toronto were respiratory in nature (approximately 95\%). This was followed by enteric outbreaks (approximately 4\%) and other types of outbreaks (less than 1\%). Note that these percentages do not add up to 100\% as the proportions in @tbl-outbreaks-by-outbreak-type were rounded to two decimal places.

@@

```{r}
#| echo: false
#| warning: false
#| label: tbl-outbreaks-by-first-known-cause
#| tbl-cap: "Outbreaks by first known cause for Toronto healthcare locations in 2023"

#### Third Table ####

@@

outbreak_cleaned_data |>
  count(`Outbreak First Known Cause`) |>
  rename(`Outbreak first known cause` = `Outbreak First Known Cause`,
         `Number of outbreaks` = n) |>
  mutate(`Proportion of total outbreaks` =
           round(`Number of outbreaks` / nrow(outbreak_cleaned_data), 2)) |>
  arrange(desc(`Number of outbreaks`)) |>
  kable()
```

@@

outbreak_cleaned_data |>
  count(`Outbreak First Known Cause`) |>
  # The reorder function used for sorting the columns in descending order was obtained
  # from Stack Overflow:
  # https://stackoverflow.com/questions/25664007/reorder-bars-in-geom-bar-ggplot2-by-value
  ggplot(mapping = aes(x = reorder(`Outbreak First Known Cause`, n), y = n)) +
  geom_col() +
  theme_bw() +
  labs(x = "Outbreak first known cause", y = "Number of outbreaks") +
  coord_flip()
```

@tbl-outbreaks-by-first-known-cause and @fig-outbreaks-by-first-known-cause illustrate that the majority of 2023 Toronto healthcare outbreaks had COVID-19 as their first known cause (approximately 66\%). This was followed by rhinovirus (approximately 7\%), respiratory syncytial virus (approximately 5\%), parainfluenza (approximately 3\%), influenza A (approximately 3\%), seasonal coronavirus (approximately 3\%), metapneumovirus (approximately 2\%), norovirus (approximately 2\%), enterovirus/rhinovirus (approximately 1\%), enterovirus (approximately 1\%), CPE (less than 1\%), neonatal Group B streptococcal disease (less than 1\%), and streptococcus pyogenes (less than 1\%). Approximately 7\% of outbreaks had a pending or unknown first known cause.

## Discussion {#sec-data-discussion}

In @sec-data-results, it was observed that long-term care homes had the greatest proportion of the total outbreaks among healthcare facilities in Toronto in 2023 (@tbl-outbreaks-by-location-type and @fig-outbreaks-by-location-type), that approximately 95\% of the outbreaks were respiratory in nature (@tbl-outbreaks-by-outbreak-type and @fig-outbreaks-by-outbreak-type), and that COVID-19 was the first known cause of a majority of the outbreaks (@tbl-outbreaks-by-first-known-cause and @fig-outbreaks-by-first-known-cause).

All three of these results share common origins. For instance, in October 2022, the Government of Ontario removed a masking requirement for caregivers and visitors in long-term care homes [@cbcltch], whose residents are primarily at or above 65 years of age with an average age of 83 years as reported in 2018 [@clri]. Since individuals aged 60 years or older are more likely to experience worse health effects after contracting COVID-19, the Public Health Agency of Canada recommends that they wear a respirator or mask to limit the transmission of this pathogen [@phaccovid]. As a result, it seems that the Government of Ontario did not heed this advice, possibly contributing to the large proportion of disease outbreaks (which include COVID-19) in Toronto long-term care homes in 2023. This appears to follow a long-standing trend in Ontario long-term care homes. For example, at the start of the COVID-19 pandemic, long-term care homes in Ontario experienced a sizable increase in COVID-19 cases [@cbcinquiry]. This was potentially made worse by the fact that not all of these facilities enforced mask-wearing and that long-term care homes were passed over for personal protective equipment in favour of hospitals [@cbcinquiry]. Moreover, the fact that several hospitals in Toronto curtailed their mask requirements in June and July of 2023 [@hospitalmasks] likely contributed to the large volume of respiratory disease outbreaks observed in @tbl-outbreaks-by-outbreak-type and @fig-outbreaks-by-outbreak-type.

@@

# Acknowledgements {#sec-acknowledgements}

Some of the R code used to create this work was adapted from @tellingstorieswithdata.

@@

>>>>

---
title: "Long-term care homes were hit hardest by 2023 disease outbreaks in Toronto healthcare facilities"
author:
  - Benny Rochwerg
thanks: "A GitHub Repository containing all data, R code, and other files used in this investigation is located here: <https://github.com/bennyrochwerg/healthcare-outbreaks-toronto>"
date: 01-23-2024
date-format: long
abstract: "In this report, 2023 Toronto Public Health data on outbreaks in healthcare facilities was analyzed. Long-term care homes comprised a majority of the outbreaks, approximately 95% of the outbreaks were respiratory, and almost two-thirds of the outbreaks had COVID-19 as their first known cause. This highlights the need for enhanced public health measures in healthcare facilities, particularly in long-term care homes. Future research should assess disease outbreaks in additional settings, such as workplaces and schools, to gain further insight into the extent of this issue."
format: pdf
number-sections: true
# The toc parameter used to create a table of contents was obtained from
# Quarto documentation:
# https://quarto.org/docs/reference/formats/pdf.html
toc: true
bibliography: references.bib
---

@@

# install.packages("tidyverse")
# install.packages("here")
library(tidyverse)
library(here)
```

@@

To do this, 2023 Toronto Public Health data on outbreaks in healthcare facilities was first obtained as described in @sec-data-overview. Based on this dataset, it was discovered that long-term care homes comprised a majority of the outbreaks, that approximately 95\% of the outbreaks were respiratory, and that almost two-thirds of the outbreaks had COVID-19 as their first known cause (@sec-data-results). As discussed in @sec-discussion, the high prevalence of outbreaks in long-term care homes has likely been exacerbated by the removal of a universal masking requirement, suggesting that the Government of Ontario should consider reinstating this policy to protect these vulnerable individuals. With respect to the structure of this paper, @sec-data includes an overview of the data and the results; @sec-discussion includes a discussion of the results; and @sec-appendix includes supplementary information.

@@

After loading the dataset using the R programming language [@citeR] and the `here` package [@citehere], the `tidyverse` [@citetidyverse] package was used to generate graphs. In doing so, R code was adapted from @tellingstorieswithdata.

@@

```{r}
#| echo: false
#| warning: false

#### First Graph Labels ####

@@

# Preparing the necessary data to create the labels
location_data <- outbreak_cleaned_data |>
  count(`Type of Location`) |>
  rename(`Type of location` = `Type of Location`,
         `Number of outbreaks` = n) |>
  mutate(`Percentage of total outbreaks` =
           round(100 * `Number of outbreaks` / nrow(outbreak_cleaned_data), 0)) |>
  arrange(desc(`Number of outbreaks`))

# Generating the labels using the paste function from the Help pane in RStudio
location_data_labels <- paste(
    location_data$`Number of outbreaks`, " (", location_data$`Percentage of total outbreaks`, "%)", sep = "")

# Adding the labels to the location data table
location_data <- location_data |>
  mutate(`Label` = location_data_labels)
```

@@

outbreak_cleaned_data |>
  count(`Type of Location`) |>
  # The reorder function used for sorting the columns in descending order was obtained
  # from Stack Overflow:
  # https://stackoverflow.com/questions/25664007/reorder-bars-in-geom-bar-ggplot2-by-value
  ggplot(mapping = aes(x = reorder(location_data$`Type of location`, location_data$`Number of outbreaks`), y = location_data$`Number of outbreaks`)) +
  geom_col() +
  theme_classic() +
  labs(x = "Type of location", y = "Number of outbreaks") +
  coord_flip() +
  # The geom_text function used to add the labels to the graph was obtained from
  # R Graphics Cookbook:
  # https://r-graphics.org/recipe-bar-graph-labels
  # The hjust argument used to align the labels was obtained from
  # Cédric Scherer's website:
  # https://www.cedricscherer.com/2021/07/05/a-quick-how-to-on-labelling-bar-graphs-in-ggplot2/
  geom_text(aes(label = location_data$`Label`), hjust = c(1, 0, 0, 0, 0, 0), colour = "magenta")
```

@fig-outbreaks-by-location-type illustrates that long-term care homes comprised a majority (approximately 56\%) of the total number of 2023 healthcare outbreaks in Toronto. This was followed by retirement homes (approximately 18\%), chronic care settings in hospitals (approximately 12\%), acute care settings in hospitals (approximately 10\%), psychiatric care settings in hospitals (approximately 2\%), and transitional care locations (approximately 1\%). Note that these percentages do not add up to 100\% as they were rounded to the nearest integer.

@@

```{r}
#| echo: false
#| warning: false

#### Second Graph Labels ####

@@

# Preparing the necessary data to create the labels
outbreak_data <- outbreak_cleaned_data |>
  count(`Type of Outbreak`) |>
  rename(`Type of outbreak` = `Type of Outbreak`,
         `Number of outbreaks` = n) |>
  mutate(`Percentage of total outbreaks` =
           round(100 * `Number of outbreaks` / nrow(outbreak_cleaned_data), 0)) |>
  arrange(desc(`Number of outbreaks`))

# Generating the labels using the paste function from the Help pane in RStudio
outbreak_data_labels <- paste(
    outbreak_data$`Number of outbreaks`, " (", outbreak_data$`Percentage of total outbreaks`, "%)", sep = "")

# Adding the labels to the outbreak data table
outbreak_data <- outbreak_data |>
  mutate(`Label` = outbreak_data_labels)
```

@@

outbreak_cleaned_data |>
  count(`Type of Outbreak`) |>
  # The reorder function used for sorting the columns in descending order was obtained
  # from Stack Overflow:
  # https://stackoverflow.com/questions/25664007/reorder-bars-in-geom-bar-ggplot2-by-value
  ggplot(mapping = aes(x = reorder(outbreak_data$`Type of outbreak`, -outbreak_data$`Number of outbreaks`), y = outbreak_data$`Number of outbreaks`)) +
  geom_col() +
  theme_classic() +
  labs(x = "Type of outbreak", y = "Number of outbreaks") +
  # The geom_text function and vjust argument used to add the labels to the
  # graph were obtained from R Graphics Cookbook:
  # https://r-graphics.org/recipe-bar-graph-labels
  geom_text(aes(label = outbreak_data$`Label`), vjust = -0.15, colour = "magenta")
```

@fig-outbreaks-by-outbreak-type shows that the vast majority of 2023 healthcare outbreaks in Toronto were respiratory in nature (approximately 95\%). This was followed by enteric outbreaks (approximately 4\%) and other types of outbreaks (less than 1\%). Note that these percentages do not add up to 100\% as they were rounded to the nearest integer.

@@

```{r}
#| echo: false
#| warning: false

#### Third Graph Labels ####

@@

# Preparing the necessary data to create the labels
first_known_cause_data <- outbreak_cleaned_data |>
  count(`Outbreak First Known Cause`) |>
  rename(`Outbreak first known cause` = `Outbreak First Known Cause`,
         `Number of outbreaks` = n) |>
  mutate(`Percentage of total outbreaks` =
           round(100 * `Number of outbreaks` / nrow(outbreak_cleaned_data), 0)) |>
  arrange(desc(`Number of outbreaks`))

# Generating the labels using the paste function from the Help pane in RStudio
first_known_cause_data_labels <- paste(" ",
    first_known_cause_data$`Number of outbreaks`, " (", first_known_cause_data$`Percentage of total outbreaks`, "%)", sep = "")

# Adding the labels to the first known cause data table
first_known_cause_data <- first_known_cause_data |>
  mutate(`Label` = first_known_cause_data_labels)
```

@@

outbreak_cleaned_data |>
  count(`Outbreak First Known Cause`) |>
  # The reorder function used for sorting the columns in descending order was obtained
  # from Stack Overflow:
  # https://stackoverflow.com/questions/25664007/reorder-bars-in-geom-bar-ggplot2-by-value
  ggplot(mapping = aes(x = reorder(first_known_cause_data$`Outbreak first known cause`, first_known_cause_data$`Number of outbreaks`), y = first_known_cause_data$`Number of outbreaks`)) +
  geom_col() +
  theme_classic() +
  labs(x = "Outbreak first known cause", y = "Number of outbreaks") +
  coord_flip() +
  # The geom_text function used to add the labels to the graph was obtained from
  # R Graphics Cookbook:
  # https://r-graphics.org/recipe-bar-graph-labels
  # The hjust argument used to align the labels was obtained from
  # Cédric Scherer's website:
  # https://www.cedricscherer.com/2021/07/05/a-quick-how-to-on-labelling-bar-graphs-in-ggplot2/
  geom_text(aes(label = first_known_cause_data$`Label`), hjust = c(1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), colour = "magenta", size = 3)
```

@fig-outbreaks-by-first-known-cause illustrates that the majority of 2023 Toronto healthcare outbreaks had COVID-19 as their first known cause (approximately 66\%). This was followed by rhinovirus (approximately 7\%), respiratory syncytial virus (approximately 5\%), parainfluenza (approximately 3\%), influenza A (approximately 3\%), seasonal coronavirus (approximately 3\%), metapneumovirus (approximately 2\%), norovirus (approximately 2\%), enterovirus/rhinovirus (approximately 1\%), enterovirus (approximately 1\%), CPE (less than 1\%), streptococcus pyogenes (less than 1\%), and neonatal Group B streptococcal disease (less than 1\%). Approximately 7\% of outbreaks had a pending or unknown first known cause.

# Discussion {#sec-discussion}

In @sec-data-results, it was observed that long-term care homes had the greatest proportion of the total outbreaks among healthcare facilities in Toronto in 2023 (@fig-outbreaks-by-location-type), that approximately 95\% of the outbreaks were respiratory in nature (@fig-outbreaks-by-outbreak-type), and that COVID-19 was the first known cause of a majority of the outbreaks (@fig-outbreaks-by-first-known-cause).

All three of these results share common origins. For instance, in October 2022, the Government of Ontario removed a masking requirement for caregivers and visitors in long-term care homes [@cbcltch], whose residents are primarily at or above 65 years of age with an average age of 83 years as reported in 2018 [@clri]. Since individuals aged 60 years or older are more likely to experience worse health effects after contracting COVID-19, the Public Health Agency of Canada recommends that they wear a respirator or mask to limit the transmission of this pathogen [@phaccovid]. As a result, it seems that the Government of Ontario did not heed this advice, possibly contributing to the large proportion of disease outbreaks (which include COVID-19) in Toronto long-term care homes in 2023. This appears to follow a long-standing trend in Ontario long-term care homes. For example, at the start of the COVID-19 pandemic, long-term care homes in Ontario experienced a sizable increase in COVID-19 cases [@cbcinquiry]. This was potentially made worse by the fact that not all of these facilities enforced mask-wearing and that long-term care homes were passed over for personal protective equipment in favour of hospitals [@cbcinquiry]. Moreover, the fact that several hospitals in Toronto curtailed their mask requirements in June and July of 2023 [@hospitalmasks] likely contributed to the large volume of respiratory disease outbreaks observed in @fig-outbreaks-by-outbreak-type.

@@

@@

## Data Cleaning {#sec-appendix-cleaning}

The data cleaning process involved filtering out some of the columns from the raw dataset and renaming some of the data entries for clarity and simplicity.
